Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-10 _tex/manuscript.tex
Built site for gh-pages

\caption{}\label{T_900f1}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{static/small-graph.png}

@@

\caption{}\label{T_4992d}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

However, for early T-category scenarios with midline extension, the
model does seem to overestimate contralateral LNL II invovlement and
underestimate the prevalence for metastases in LNL III. This likely
stems from the the small sample size of this rare scenario as hinted at
by the wide Beta posteriors.

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-with-ipsi.png}

@@

\begin{itemize}
\tightlist
\item
  compute risks for occult disease, given diagnosis
\end{itemize}

\subsection{Distributions over States}\label{distributions-over-states}

\includegraphics{manuscript_files/figure-pdf/cell-6-output-1.pdf}

@@

\section{Appendix}\label{appendix}

@@

\caption{}\label{T_07ae0}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

>>>>

\caption{}\label{T_c4fec}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{static/small-graph.png}

@@

\caption{}\label{T_b241f}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

However, for early T-category scenarios with midline extension, the
model does seem to overestimate contralateral LNL II and III
invovlement. This likely stems from the the small sample size of this
relatively rare scenario as hinted at by the wide beta posteriors.

@@

\includegraphics{figures/fig-model-prevalences-with-ipsi.png}

@@

In this section, we investigate how the model may be applied clinically:
We want to estimate the risk for occult disease in some or all LNLs,
given the patient's individual diagnosis. In terms of our model, this
diagnosis consists of the T-category, the lateralization of the tumor
(does it extend over the mid-sagittal line?), and which LNLs are
clinically involved, e.g.~because some lymph nodes appear enlarged in an
MRI scan or show increased glycolytic activity on an FDG PET/CT scan.

\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-risks.png}

\caption{\label{fig-model-risks}Histograms over the predicted risk of
occult contralateral LNL II invovlement, shown for some combinations of
T-category, tumor lateralization, and ipsilateral clinical involvement.
The contralateral side was always assumed to be clinically negative.}

Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
in contralateral LNL II for four interesting combinations of these three
risk factors. This risk is computed \emph{given} a CT diagnosis
assessing the per-LNL clinical involvement for which we assumed a
sensitivity of 81\% and a specificity of 76\%
\citep{debondt_detection_2007}.

The variable impacting the prediction for contralateral involvement in
our model is the tumor's lateralization. For example, a patient with a
clearly lateralized early T-category tumor and a clinically N0 neck is
assigned a 1-2\% risk for occult disease in contralateral LNL II. Under
the same scenario, but \emph{with} mid-sagittal extension, the risk
jumps to almost 7\%.

T-category plays a lesser role: Considering the scenario of a tumor that
crosses the mid-sagittal line and an ipsilateral neck where at least LNL
III is clinically involved, the risk for occult contralateral LNL II
disease is around 8.5\%. For the same scenario and an advanced
T-category tumor, the risk increases to 11\%.

Lastly, the predicted risk is also correlated via the time-steps to the
degree of ipsilateral involvement. Changing the aforementioned scenario
(advanced T-category, midline extension, ipsilateral LNL III clinically
involved) to one where the patient presents with a clinically N0 neck,
the risk for occult disease in the contralateral LNL II falls from 11\%
to 9.5\%.

Taken together, T-category and ipsilateral involvement may still
considerably impact the risk prediction for contralateral involvement:
In figure~\ref{fig-model-risks} the scenarios underlying the orange
(6.5\% ) and the red (11\% ) histograms differ in T-category (early vs
advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III
involved).

@@

\section{Contralateral Prevalence of
Involvement}\label{contralateral-prevalence-of-involvement}

@@

\caption{}\label{T_bb1c4}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

\section{Marginalizing Beta Posteriors over
Unknowns}\label{marginalizing-beta-posteriors-over-unknowns}

\begin{itemize}
\tightlist
\item[$\square$]
  write short theory section about why and how to compute the posteriors
  if some of the data (here midline extension) is unknown
\end{itemize}
